On the Uncertainty of Wind Power Forecasts

### 2019-06-17 LaTeX_code/poster/latex/optimization_project.tex
updated code after finalizing the poster. working optmization, full data

%%%%%%%%% TITLE
\title{ On the Uncertainty of Wind Power Forecasts \\  \small{(AMCS Optimization course final project submission)}}

@@

    We are able to simulate and quantify uncertainties in a variety of wind power generation forecasts while taking into account the skewness of the errors. This method is non-intrusive and is independent of forecasting technology or future developments. Through optimization, we update and tune the parameters of our SDE as we receive new sets of observations and their associated forecasts. Additionally, we are able to compare in a quantitative manner the different forecast technologies and how they behave in different real-world scenarios. This model is to be extended to the uncertainty quantification of other power generation sources such as the uncertainties in the power generation of solar power plants. This introduces new challenges in terms of optimization and modeling. Finally, we apply our model to synthetic and real wind power generation data for benchmarking. We apply the model to Uruguayan wind power production as determined by historical data and corresponding numerical forecasts for the full year 2016-2017.

%-------------------------------------------------------------------------
\subsection{Model}

We propose the following parameterized stochastic differential equation (SDE), 
\begin{equation}
\begin{split}
dV_t &= a(V_t; \bm{\theta}) dt + b (V_t; \bm{\theta} ) dW_t \quad t > 0 \\
V_0 & = v_0
\end{split}\label{main}
\end{equation}

\begin{itemize}
\item $\Delta_N$: any function of the number of samples, $N$. For example, $\Delta_N = \frac{1}{N}$.
\item $a(X_t; \bm{\theta})$: Drift function 
\item $b (X_t; \bm{\theta} )$: Diffusion function 
\item $\bm{\theta}$: a vector of parameters
\item $dW_t$: Standard Wiener random process.
\end{itemize}

To respect the physical restriction of this model, we consider the following diffusion and drift functions for a parameter vector $\bm{\theta} = (\theta, \alpha)$,  
\begin{equation}
\begin{split}
a(V_t; \bm{\theta}) &= - \theta V_t \\
b(V_t; \bm{\theta}) &=\sqrt{2 \theta \alpha p_t(1-p_t) (V_t +p_t ) (1-V_t-p_t)}  \\
\end{split}
\end{equation}

Where
\begin{itemize}
\item $p_t$: Numerical wind power generation forecast.
\item $\theta >0$: Mean reversion parameter.
\item $\alpha>0$: Variablity parameter.
\end{itemize}

The SDE in (\ref{main}) define the stochastic process $V_t$ which can be characterized by its transition denisity. Consider a set of M paths with N observations each, $ V^{M,N}=\{ V_{t_1^{M,N}} , V_{t_2^{M,N}} ,\ldots , V_{t_N^{M,N}} \}$ observed within intervals $\Delta_N$. Then, the transitions of $V_t$ are given by the likelihood function,

@@

%-------------------------------------------------------------------------
\subsection{Results}
 We are able to obtain the parameters based on the complete data sets mentioned earlier. The parameters are given by $(\theta^*, \alpha^*)= (12,2)$. We show in Figure(\ref{fig:ISO}) the contours of the log-likelihood in $(\ref{Likelihood})$ using the full data set.

@@

In Figure(\ref{fig:6hr}) and (\ref{fig:72hr}), we simulate sample forecast errors using the optimal parameters  $(\theta^*, \alpha^*)= (12,2)$ and obtain the associated confidence bands empirically using 10,000 sample paths for each forecast. Note that the confidence bands obtained are non-trivial and reflect the complexity of the uncertainties in the numerical wind power generation forecasts.

@@

\begin{figure}[t]
\begin{center}
%\fbox{\rule{0pt}{2in} \rule{0.9\linewidth}{0pt}}
   \includegraphics[width=0.9\linewidth]{ISO_lines_detailed.pdf}
   \includegraphics[width=0.9\linewidth]{ISO_lines_zoom_2.pdf}
\end{center}
   \caption{The Beta distribution log-likelihood evaluated given the complete data set of 1217 sample paths. The optimal parameter found are indicated by the blue dot.}
\label{fig:ISO}
\end{figure}

>>>>

%%%%%%%%% TITLE
\title{ On the Uncertainty of Wind Power Forecasts }  % \\  \small{Report}}

@@

    We are able to simulate and quantify uncertainties in a variety of wind power generation forecasts while taking into account the skewness of the errors. This method is non-intrusive and is independent of forecasting technology or future developments. Through optimization, we update and tune the parameters of our SDE as we receive new sets of observations and their associated forecasts. Additionally, we are able to compare in a quantitative manner the different forecast technologies and how they behave in different real-world scenarios. This model is to be extended to the uncertainty quantification of other power generation sources such as the uncertainties in the power generation of solar power plants. This introduces new challenges in terms of optimization and modeling. Finally, we apply our model to synthetic and real wind power generation data for benchmarking. We apply the model to Uruguayan wind power production as determined by historical data and corresponding numerical forecasts for the year 2016-2017.

%-------------------------------------------------------------------------
\section{Model}

\subsection{Physical Model}

Let $X_t$ be the  wind power generation forecasts stochastic process defined by the  following parameterized stochastic differential equation (SDE), 
\begin{equation}
\begin{split}
dX_t &= a(X_t; \bm{\theta}) dt + b (X_t; \bm{\theta} ) dW_t \quad t > 0 \\
X_0 & = X_0
\end{split}\label{main}
\end{equation}

\begin{itemize}
%\item $\Delta_N$: any function of the number of samples, $N$. For example, $\Delta_N = \frac{1}{N}$.
\item $a(X_t; \bm{\theta})$: Drift function 
\item $b (X_t; \bm{\theta} )$: Diffusion function 
\item $\bm{\theta}$: a vector of parameters
\item $dW_t$: Standard Wiener random process.
\end{itemize}

%\subsection{Base Model}
%
%Let $V_t$ be the deviation of the wind power generation forecasts from  actual wind power generation. That is, $V_t$ models the errors or uncertainty of a given set of wind power generation forecasts. 
%
%We propose a forecast-error model based on the following parameterized stochastic differential equation (SDE), 
%\begin{equation}
%\begin{split}
%dV_t &= a(V_t; \bm{\theta}) dt + b (V_t; \bm{\theta} ) dW_t \quad t > 0 \\
%V_0 & = v_0
%\end{split}\label{main}
%\end{equation}
%
%
%\begin{itemize}
%%\item $\Delta_N$: any function of the number of samples, $N$. For example, $\Delta_N = \frac{1}{N}$.
%\item $a(V_t; \bm{\theta})$: Drift function 
%\item $b (V_t; \bm{\theta} )$: Diffusion function 
%\item $\bm{\theta}$: a vector of parameters
%\item $dW_t$: Standard Wiener random process.
%\end{itemize}

\subsection{Physical Restrictions}
It is necessary to keep  the power generation forecast plus its errors  inside the range $[0,1]$ after normalizing the power generated to the capacity of the power plant.  To ensure that, we choose our model to have zero diffusion near the boundaries. Hence, we  formulate the following diffusion  term, 
\begin{equation*}
\begin{split}
b(V_t; \bm{\theta}) &=  \sqrt{2 \theta \alpha p_t(1-p_t) X_t (1-X_t)}    \\
&= \sqrt{2 \theta \alpha p_t(1-p_t) (V_t +p_t ) (1-V_t-p_t)}  \\
\end{split}
\end{equation*}
It remains to control the drift term. As a first requirement, we need the drift term to be mean reverting to the power generation forecast. Thus, we formulate the following drift, 
\begin{equation}
a(V_t; \bm{\theta}) =  - \theta (X_t - p_t)
\end{equation}
However, this does not track the rate of change of the forecast $\dot{p}$ which  results in $X_t$ being out of sync with the forecast  as can be seen in Figure (\ref{sine_wave_shift})

\begin{figure}[t]
\begin{center}
%\fbox{\rule{0pt}{2in} \rule{0.9\linewidth}{0pt}}
   \includegraphics[width=0.8\linewidth]{conv_sine_shift.pdf}
\end{center}
   \caption{Example of the process $X_t$ following a sine wave without derivative tracking.}
\label{fig:long}
\label{fig:onecol}
\label{sine_wave_shift}
\end{figure}

For clarity of exposition, we switch to the point of view of the forecast prediction process $X_t = V_t + p_t$ and we include the derivative tracking in the base model,
\begin{equation}
\begin{split}
dX_t &= \hat{a}(X_t; \bm{\theta}) dt + \hat{b} (X_t; \bm{\theta} ) dW_t \quad t > 0 \\
X_0 & = x_0
\end{split}\label{main_X}
\end{equation}
where 
\begin{equation}
\begin{split}
\hat{a}(X_t; \bm{\theta}) &= \dot{p}_t + a(X_t; \bm{\theta}) =  \dot{p}_t - \theta \left(X_t - p_t\right)   \\
\hat{b}(X_t; \bm{\theta}) &=  b(X_t; \bm{\theta}) =  \sqrt{2 \theta \alpha p_t(1-p_t) X_t (1-X_t)}    \\
\end{split}
\end{equation}
Here the term $\dot{p}_t $ is not controlled to maintain that $X_t$ stays inside the range $[0,1]$. In other words, the zero drift line defined by  setting $\hat{a}(X_t; \bm{\theta}) =0$ must  be contained inside the range $[0,1]$. This line is given by   $ \frac{\dot{p}_t}{\theta}  + p_t $, we call it the line of zero drift or  the line of mean stationarity.
Thus, must have that, 
\begin{equation}
 0 \leq  \frac{\dot{p}_t}{\theta}  + p_t  \leq 1
\end{equation}
to keep the process $X_t$ inside the interval $[0,1]$.
The condition  can be rewritten as, 
\begin{equation}
\frac{- |\dot{p_t}|}{p_t} \leq \theta \leq \frac{|\dot{p}_t|}{1- p_t}
\end{equation}

To enforce the condition, we  define an adjusted drift $\theta = \theta_0 f( p_t, \dot{p}_t) $  which satisfies the condition when
\begin{equation}
\theta = \theta_0 f( p_t, \dot{p}_t) \leq \frac{|\dot{p}_t|}{\min (p_t, 1-p_t)} 
\end{equation}

Hence, we choose $\theta$ such that,
\begin{equation}
\theta = \max \left( \theta_0 \ , \ \frac{|\dot{p}_t|}{\min (p_t, 1-p_t)}  \right )
\end{equation}
\subsection{Model prescription }
We conclude that in order to respect the physical restrictions discussed above, we prescribe the following specifications and the final model becomes,
\begin{equation}
\begin{split}
dV_t &= a(V_t; \bm{\theta}) dt + b (V_t; \bm{\theta} ) dW_t \quad t > 0 \\
V_0 & = v_0
\end{split}\label{main}
\end{equation}

\begin{equation}
\begin{split}
a(V_t; \bm{\theta}) &= - \theta V_t \\
b(V_t; \bm{\theta}) &=\sqrt{2 \theta \alpha p_t(1-p_t) (V_t +p_t ) (1-V_t-p_t)}  \\
\end{split}
\end{equation}

with
\begin{equation}
\theta = \max \left( \theta_0 \ , \ \frac{|\dot{p}_t|}{\min (p_t, 1-p_t)}  \right )
\end{equation}

Where 
\begin{itemize}
\item $p_t$: Numerical wind power generation forecast.
\item $\theta >0$: Mean reversion parameter.
\item $\alpha>0$: Variability parameter.
\end{itemize}

The  SDE above defines the stochastic process $V_t$ which can be characterized by its transition denisity. Consider a set of M paths with N observations each, $ V^{M,N}=\{ V_{t_1^{M,N}} , V_{t_2^{M,N}} ,\ldots , V_{t_N^{M,N}} \}$ observed in intervals of $\Delta_N$. Then, the transitions of $V_t$ are given by the likelihood function,

@@

%-------------------------------------------------------------------------
\subsection{Results}
 We are able to obtain the parameters based on the complete data sets mentioned earlier. The parameters are given by $(\theta^*, \alpha^*)\approx (8,1)$. We show in Figure(\ref{fig:ISO}) the contours of the log-likelihood in $(\ref{Likelihood})$ using the full data set.

@@

In Figure(\ref{fig:6hr}) and (\ref{fig:72hr}), we simulate sample forecast errors using the optimal parameters  $(\theta^*, \alpha^*)\approx (8,1)$ and obtain the associated confidence bands empirically using 10,000 sample paths for each forecast. Note that the confidence bands obtained are non-trivial and reflect the complexity of the uncertainties in the numerical wind power generation forecasts.

@@

%\begin{figure}[t]
%\begin{center}
%%\fbox{\rule{0pt}{2in} \rule{0.9\linewidth}{0pt}}
%   \includegraphics[width=0.9\linewidth]{ISO_lines_detailed.pdf}
%   \includegraphics[width=0.9\linewidth]{ISO_lines_zoom_2.pdf}
%\end{center}
%   \caption{The Beta distribution log-likelihood evaluated given the complete data set of 1217 sample paths. The optimal parameter found are indicated by the blue dot.}
%\label{fig:ISO}
%\end{figure}

\begin{figure}[t]
\begin{center}
%\fbox{\rule{0pt}{2in} \rule{0.9\linewidth}{0pt}}
   \includegraphics[width=0.5\linewidth]{ellipse100_samples_dN=14e-02.pdf}
   \includegraphics[width=0.5\linewidth]{ellipse500_samples_dN=14e-02.pdf}
   \includegraphics[width=0.5\linewidth]{ellipse1193_samples_dN=14e-02.pdf}
\end{center}
   \caption{Shrinkage of the ellipse determined by the Hessian matrix of the log-likelihood around the point of optimality $(\theta, \alpha)\approx (8,1)$}
\label{ellipse_drawing}
\end{figure}

 \begin{figure}[t]
\begin{center}
%\fbox{\rule{0pt}{2in} \rule{0.9\linewidth}{0pt}}
   \includegraphics[width=0.5\linewidth]{ISO_100_samples_dN=14e-02.pdf}
   \includegraphics[width=0.5\linewidth]{ISO_500_samples_dN=14e-02.pdf}
   \includegraphics[width=0.5\linewidth]{ISO_1193_samples_dN=14e-02.pdf}
\end{center}
   \caption{Contour plot of the log-likelihood as more forecasts and data is added, point of optimality $(\theta, \alpha)\approx (8,1)$ indicated in black.}
\label{contour}
\end{figure}

\begin{figure}[t]
\begin{center}
%\fbox{\rule{0pt}{2in} \rule{0.9\linewidth}{0pt}}
   \includegraphics[width=0.5\linewidth]{Eigen_conv_samples_dN=14e-02.pdf}
   \includegraphics[width=0.5\linewidth]{ellipse_conv_samples_dN=14e-02.pdf}
\end{center}
   \caption{Convergence of the major and semi-axis of the ellipse  of the Hessian of the log-likelihood at the point optimality $(\theta, \alpha)\approx (8,1)$. Note that it is slightly faster than the expected rate of $1/\sqrt{M}$. This due to the fact that segments of some of the $M$ are uncorrelated, thus acting as an additional  uncorrelated sample to the rest of the path. }
\label{Ellipse_convergance}
\end{figure}
